The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-11 paper.tex
more from manuscript structure

Uberon, the Uber Anatomy Ontology~\citep{uberon}...
BSPO~\citep{bspo}...
NCIT~\citep{ncit}...
The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)

@@

To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.

%\url{https://bioportal.bioontology.org/ontologies/FMA}

>>>>

Uberon, the Uber Anatomy Ontology~\citep{uberon}...
BSPO~\citep{bspo}...
NCIT~\citep{ncit}...
The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
Read Codes Version 3 (RCD)~\citep{rcd}...

@@

To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.
%
The main requirement for the intended ontology to cover the defined tasks is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}.
Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
The FMA, for example, does not have single equivalent concepts to represent bone compounds or bone parts, see \cref{sec:annodc}.
%(see section 3.2.2).
Several partial subtrees would have to be combined for this purpose.
Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.
None of the evaluated ontologies could cover all defined requirements and adequately support the specific anthropological use cases.
We decided to develop a new ontology because the integration (extension and adaptation) of all necessary parts from different ontologies would have been much more complex.
